---
title: "Introduction to EpidigiR"
author: "Esther Atsabina Wanjala"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to EpidigiR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## title: "Introduction to EpidigiR: Epidemiological Analysis and Visualization" output: rmarkdown::html_vignette vignette: \> %\\VignetteIndexEntry{Introduction to EpidigiR} %\\VignetteEngine{knitr::rmarkdown} %\\VignetteEncoding{UTF-8}

# Introduction to EpidigiR

EpidigiR is an R package for epidemiological analysis, modeling, and visualization, designed with minimal dependencies and comprehensive functionality. It provides three main functions to cover 12 epidemiological topics:

-   **epi_analyze**: Performs summary statistics, SIR modeling, DALY calculations, age standardization, diagnostic test evaluation, and NLP keyword extraction.
-   **epi_model**: Handles clinical trial power calculation, survival analysis, SNP association, logistic regression, k-means clustering, Random Forest, and SVM.
-   **epi_visualize**: Creates visualizations for prevalence mapping, epidemic curves, scatter plots, and boxplots.

The package includes nine datasets to support these analyses: epi_prevalence, sir_data, geno_data, ml_data, nlp_data, clinical_data, daly_data, survey_data, diagnostic_data, and survival_data.

This vignette demonstrates how to use these functions and datasets for various epidemiological tasks.

## Setup

```{r setup, include=FALSE}
    knitr::opts_chunk$set(echo = TRUE)

# Install required dependencies if not already installed
required_packages <- c("deSolve", "sp", "tm", "glmnet", "caret", "kernlab", "survival")
for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
}
# Load required packages
library(deSolve)
library(sp)
library(tm)
library(glmnet)
library(caret)
library(kernlab)
library(survival)
library(randomForest)
library(EpidigiR)
# Ensure ml_data and clinical_data outcomes are factors for caret models
data(ml_data)
ml_data$outcome <- as.factor(ml_data$outcome)
data(clinical_data)
clinical_data$outcome <- as.factor(clinical_data$outcome)
```

## Datasets

The package includes the following datasets:

-   **epi_prevalence**: Disease prevalence by region and age group, with spatial coordinates (12 rows).
-   **sir_data**: Simulated SIR model output (50 rows).
-   **geno_data**: Genotype and case-control data for SNP analysis (100 rows).
-   **ml_data**: Patient data for machine learning (logistic regression, clustering, Random Forest, SVM; 100 rows).
-   **nlp_data**: Epidemiological text data for NLP (100 rows).
-   **clinical_data**: Clinical trial data for power calculations and outcome analysis (200 rows).
-   **daly_data**: Data for DALY calculations (20 rows).
-   **survey_data**: Data for age standardization (20 rows).
-   **diagnostic_data**: Data for diagnostic test evaluation (10 rows).
-   **survival_data**: Data for survival analysis (100 rows).

## Examples

### 1. Summary Statistics

Calculate incidence and prevalence using epi_prevalence.

```{r}
data(epi_prevalence)
epi_analyze(epi_prevalence, outcome = "cases", group = "region", type = "summary")
```

### 2. SIR Epidemic Model

Simulate an SIR model and visualize the epidemic curve.

```{r}
sir_result <- epi_analyze(data = NULL, outcome = NULL, type = "sir", N = 1000, beta = 0.3, gamma = 0.1, days = 50)
    epi_visualize(sir_result, x = "time", y = "Infected", type = "curve", main = "Epidemic Curve")
```

### 3. Map Disease Prevalence

Visualize prevalence by region using epi_prevalence.

```{r}
data(epi_prevalence)
coordinates(epi_prevalence) <- ~lon+lat
epi_visualize(epi_prevalence, x = "prevalence", type = "map", main = "Prevalence Map")
```

### 4. Clinical Trials (Power Calculation)

Logistic Regression with Clinical Data.

```{r}
data(clinical_data)
    clinical_data$outcome <- as.factor(clinical_data$outcome)
    model <- epi_model(clinical_data, formula = outcome ~ age + health_score + dose, type = "logistic")
    head(model$predictions)
```

## Random Forest with Clinical Data

```{r}
rf_model <- epi_model(clinical_data, formula = outcome ~ age + health_score + dose, type = "rf")
    head(rf_model$predictions)

```

### 5. Clinical Trials (Outcome Analysis)

Analyze treatment outcomes using logistic regression.

```{r}
data(clinical_data)
epi_model(clinical_data, formula = outcome ~ age + health_score + dose, type = "logistic")
```

### 6. Global Health Burden (DALY)

Calculate DALYs using daly_data.

```{r}
data(daly_data)
epi_analyze(daly_data, outcome = NULL, type = "daly")
```

### 7. Genomic and Data Integration (SNP Association)

Perform SNP association analysis using geno_data.

```{r}
data(geno_data)
epi_model(geno_data, formula = outcome ~ snp1 + snp2, type = "snp")

```

### 8. Survey and Demographics (Age Standardization)

Perform age standardization using survey_data.

```{r}
data(survey_data)
epi_analyze(survey_data, outcome = NULL, type = "age_standardize")
```

### 9. Survey and Demographics (Logistic Regression)

Predict disease risk using ml_data.

```{r}
data(ml_data)
epi_model(ml_data, formula = outcome ~ age + exposure + genetic_risk, type = "logistic")
```

### 10. Survival Analysis

Perform survival analysis using survival_data.

```{r}
data(survival_data)
epi_model(survival_data, type = "survival")
```

### 11. NLP Tools

Extract keywords from epidemiological texts using nlp_data.

```{r}
data(nlp_data)
    nlp_result <- epi_analyze(nlp_data, outcome = NULL, type = "nlp")
    nlp_result$keywords
```

### 12. Machine Learning and AI (K-means Clustering)

Cluster patients using ml_data.

```{r}
data(ml_data)
epi_model(ml_data[, c("age", "exposure", "genetic_risk")], type = "kmeans", k = 3)
```

### 13. Machine Learning and AI (Random Forest)

Train a Random Forest model using ml_data.

```{r}
data(ml_data)
epi_model(ml_data, formula = outcome ~ age + exposure + genetic_risk, type = "rf")
```

### 14. Machine Learning and AI (SVM)

Train an SVM model using ml_data.

```{r}
data(ml_data)
    ml_data$outcome <- as.factor(ml_data$outcome)
    svm_model <- epi_model(ml_data, formula = outcome ~ age + exposure + genetic_risk, type = "svmRadial")
    svm_model$performance
```

### 15. Evaluation of Diagnostic Tests

Evaluate diagnostic test performance using diagnostic_data.

```{r}
data(diagnostic_data)
epi_analyze(diagnostic_data, outcome = NULL, type = "diagnostic")
```

## Visualization Examples

Visualize clinical trial outcomes by treatment arm.

```{r}
data(clinical_data)
epi_visualize(clinical_data, x = "arm", y = "outcome", type = "boxplot", main = "Outcome by Treatment Arm")
```

Visualize machine learning data (age vs. outcome).

```{r}
data(ml_data)
epi_visualize(ml_data, x = "age", y = "outcome", type = "scatter", main = "Age vs. Disease Outcome")
```

## Conclusion

EpidigiR provides a lightweight, comprehensive toolkit for epidemiological analysis, with three functions and nine datasets covering all major topics. Use epi_analyze, epi_model, and epi_visualize to perform and visualize a wide range of analyses, from SIR modeling to advanced machine learning with Random Forest and SVM.

## License

EpidigiR is released under the MIT License © 2025 Esther Atsabina Wanjala.
